How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-06-12 doc/paper.tex
trying to understand data

@@

These temperature estimates are measured from benthic forams, and is an estimate of deep water temperature. The organisms included in this study are all planktonic, so these temperature estimates do not necessarily represent the actual environment these organisms inhabit. However, detailed ocean surface temperature estimates do not exist for the entirety of the Cenozoic. So, while deepwater temperature estimates are not ideal for understanding the extinction risk of planktonic taxa, we interpret our estimate of temperature as reflecting global climate state as opposed to a taxon-specific ecological descriptor.

@@

We always included species age at time of observation (i.e. observed prior duration) as a non-nested varying intercept term. This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.

@@

In-sample forecasting is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data. This is a type of sensitivity analysis for understanding the quality of model fit to the data, and how well our model predicts extinction probability for observations that we used to fit out model initially. In-sample forecasting measures, however, are not necessarily good estimates of the model's ability to predict data from the future \citep{ESL}.

@@

\textit{k}-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016}. First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. Continuing, the model is then fit to the first three folds combined and is then used to forecast extinction probabilities for the fourth fold. Next, the model is fit to the first four folds combined and then is used to forecast the fifth fold. This process continues until $k - 1$ folds are included in the fitting the model and the final fold is predicted from this model. When combined, the results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.

Some readers may be familiar with metrics such as AIC, WAIC, or LOOIC for comparing model fit. These metrics are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL}. Because we've estimated out-of-sample performance using cross-validation, these measures would be redundant, and in the case on single value measures such as AIC, much less informative than an entire posterior distributions. When out-of-sample estimates from a complex model are noticeably worse than out-of-sample estimates from a simpler model, than this is evidence for the complex model being overfit to the data \citep{ESL}. In contrast a high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data).

@@

% ROC model comparison
\begin{figure}[ht]
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
  \caption{In-sample}
  \label{fig:auc_hist}
 \end{subfigure}
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
  \caption{Out-of-sample}
  \label{fig:fold_auc}
 \end{subfigure}
 \caption{Comparisons of measures of model performance for both in-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) cross-validation. We used area under the receiver operating charactered curve (AUC) estimates for each of our four models as our measure of performance. These estimates are calculated from the models posterior predictive distribution or from predictions made to new data, respectively. Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in our dataset. Models with higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1. See Table \ref{tab:model_def} for an explanation of the four models (C, V, CP, VP).}
 \label{fig:auc_compare}
\end{figure}

@@

\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
 \caption{Understanding model adequacy over time and taxonomic group by comparing in-sample forecasting performance measured by AUC for each of the four models. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models.}
 \label{fig:auc_taxon_time}
\end{figure}

@@

In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts. The major exception to this pattern are our estimates for the diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random. The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
 \caption{Comparison of our models ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models.}
 \label{fig:fold_auc_taxon_time}
\end{figure}

@@

\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
 \caption{Comparing our models' ability to forecast out-of-sample extinction compared to our models' in-sample forecasts. Congruence between these measures indicates that our models are not necessarily overfit to the data. This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}. A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast. See Table \ref{tab:model_def} for a description of each of the four models.}
 \label{fig:oos_ins_diff}
\end{figure}

@@

To implement our VP model in \begin{texttt}rstanarm\end{texttt}, where ``data'' is a data.frame object of all necessary data (response, covariates), is coded as:
\begin{verbatim}
form <- event ~ range + range_diff + temp + temp_lag + 
        (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
        (1 | age/phylum), 
stan_glmer(formula = form,
      data = data, 
      family = 'binomial',
      prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
      prior_intercept = normal(-2, 5, autoscale = FALSE), 
      prior_aux = cauchy(0, 1, autoscale = FALSE), 
      chains = 4,
      thin = 4,
      adapt_delta = 0.999999)
\end{verbatim}

Similarly, our VP model can be implemented using the \begin{texttt}brms\end{texttt} Stan interface \citep{brms2017,brms2018} as:
\begin{verbatim}
priors <- c(set_prior('normal(-2, 5)', class= 'Intercept'),
      set_prior('normal(0, 1)', class = 'b'),
      set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
      set_prior('cauchy(0, 1)', class = 'sd'),
      set_prior('lkj(1)', class = 'cor'))
form <- bf(event ~ range + range_diff + temp + temp_lag +
      (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
      (1 | age/phylum))
brmfit <- brm(formula = form,
       data = data, 
       family = bernoulli(), 
       prior = priors,
       chains = 4, 
       thin = 4,
       control = list(adapt_delta = 0.999999)
\end{verbatim}

>>>>

We also included how a taxon's geographic range has changed since its last three observation times. We measured this change in geographic range by calculating the difference in geographic range between an observation and that taxon's three previous occurrences. If there are not enough previous observations to calculate, then that value is recorded as a 0. These differences were calculated after minimum spanning tree distance was transformed and standardized (see Section \ref{sec:data_desc}).

@@

We also included the global temperature from the previous time that taxon was observed. If there are not enough previous observations to calculate, then that value is recorded as a 0. This lag was calculated after global temperature was transformed and standardized (see Section \ref{sec:data_desc}).

Mg/Ca based temperature estimates are measured from benthic forams, and are an estimate of deep water ocean temperature. The organisms in this study are all planktonic, Mg/Ca based temperature estimates do not describe the exact environment these organisms inhabit. Ideally, we would have detailed ocean surface water temperature estimates for the entire globe for the entire Cenozoic. Unfortunately, that type of data does not exist. So, we interpret our temperature estimates as reflecting the global climate state that an organisms experiences, and not as a descriptor of that taxon's local environmental ecology.

@@

We always included species age at time of observation (i.e. observed prior duration) as a varying-intercept term. This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}. The effect of species age is allowed to vary by taxonomic group.

Similarly, we included time of obseration as an additional varying-intercept term to account for changes average global extinction risk over time that are not related to the covariates included in this model. This varying-intercept is further allowed to vary by taxonomic group. This varying-intercept term allows us to tease apart the differences in extinction risk associated with time of observation versus age since first observation. An important note is that for our V and VP models, the covariation between this varying-intercept and the varying-slopes of our covariates is explicitly modeled (see Section \ref{sec:model_desc}).

@@

In-sample forecasting is a posterior predictive check in that we are estimating our model's ability to recapitulate the data to which it was fit. Posterior predictive checks are a type of sensitivity analysis because we are checking the quality of model's fit to the data. If our models have poor in-sample forecasting performance, than our models are not adequate descriptors of the data and will most likely make poor out-of-sample predictions. In-sample forecasting measures, however, are not the same as understanding our models' ability to forecast future extinctions or if our models are overfit to our data and produce biased out-of-sample estimates \citep{ESL}.

@@

\textit{k}-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016,ESL}. Prior to cross-valdiation, the data is divided into \textit{k} nearly even segments or folds -- for a time series, this means the data is divided into \textit{k} continuous sequences. Next, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. Continuing, the model is then fit to the first three folds combined and is then used to forecast extinction probabilities for the fourth fold. Next, the model is fit to the first four folds combined and then is used to forecast the fifth fold. This process continues until $k - 1$ folds are included in the fitting the model and the final fold is predicted from this model. When combined, the results from these forecasts are then combined to yield our estimate of expected out-of-sample performance. In 5-fold cross validaition, the data is divided into five folds the cross-validation procedure yields predictions for four of the folds.

Cross-validation is a procedure for estimating a models expected out-of-sample error, which is a similar goal to metrics such as AIC or WAIC \citep{ESL,Gelman2013d}. Measures like AIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}. Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, then out-of-sample predictions will be very biased and inaccurate \citep{ESL}. Similarly, a high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data). Cross-validation is preferable to simple metrics such as AIC because instead of a single value, we have an entire posterior distribution of estimates.

@@

% ROC model comparison
\begin{figure}[ht]
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
  \caption{In-sample}
  \label{fig:auc_hist}
 \end{subfigure}
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
  \caption{Out-of-sample}
  \label{fig:fold_auc}
 \end{subfigure}
 \caption{Comparisons of measures of model performance for both in-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) cross-validation. The area under the receiver operating charactered curve (AUC) was calculated for each model. these estimates are calculated from the models posterior predictive distribution (\ref{fig:auc_hist}) or from predictions made to new data (\ref{fig:fold_auc}), respectively. Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in our dataset. Models with higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1. See Table \ref{tab:model_def} for an explanation of the four models (C, V, CP, VP).}
 \label{fig:auc_compare}
\end{figure}

@@

\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
 \caption{Understanding model adequacy over time and taxonomic group by comparing in-sample forecasting performance measured by AUC for each of the four models. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:auc_taxon_time}
\end{figure}

@@

In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts. The major exception to this pattern are our estimates for the diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random. The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
 \caption{Comparison of our models ability to forecast future extinction events as measured by out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:fold_auc_taxon_time}
\end{figure}

@@

\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
 \caption{Comparing our models' ability to forecast out-of-sample extinction compared to our models' in-sample forecasts. Congruence between these measures indicates that our models are not necessarily overfit to the data. This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}. A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast. See Table \ref{tab:model_def} for a description of each of the four models (C, V, CP, VP).}
 \label{fig:oos_ins_diff}
\end{figure}

@@

To implement our VP model in \begin{texttt}rstanarm\end{texttt}, where ``data'' is a data.frame object of all necessary data (response, covariates), is coded as:
\begin{verbatim}
form <- event ~ 
        range + range_diff1 + range_diff2 + range_diff3 + 
        temp + temp_lag + 
        (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
        (1 | age/phylum), 
stan_glmer(formula = form,
      data = data, 
      family = 'binomial',
      prior = normal(c(-1, 0, 0, 0, 0, 0), rep(1, 6), autoscale = FALSE), 
      prior_intercept = normal(-2, 5, autoscale = FALSE), 
      prior_aux = cauchy(0, 1, autoscale = FALSE), 
      chains = 4,
      thin = 4,
      adapt_delta = 0.999999)
\end{verbatim}

Similarly, our VP model can be implemented using the \begin{texttt}brms\end{texttt} Stan interface \citep{brms2017,brms2018} as:
\begin{verbatim}
priors <- c(set_prior('normal(-2, 5)', class= 'Intercept'),
      set_prior('normal(0, 1)', class = 'b'),
      set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
      set_prior('cauchy(0, 1)', class = 'sd'),
      set_prior('lkj(1)', class = 'cor'))
form <- bf(event ~ 
           range + range_diff1 + range_diff2 + range_diff3 + 
           temp + temp_lag +
           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
           (1 | age/phylum))
brmfit <- brm(formula = form,
       data = data, 
       family = bernoulli(), 
       prior = priors,
       chains = 4, 
       thin = 4,
       control = list(adapt_delta = 0.999999)
\end{verbatim}
